### 2017-02-27 2017mergingmerges/section/04-Fusion.tex
Rename XX-Fusion.tex to 04-Fusion.tex
Shift discussion of fusion order to Fusion In Practice section, and elaborate

\clearpage{}
\section{Fusion}
\label{s:Fusion}

The core fusion algorithm works by constructing a static schedule for a pair of processes, which is repeated over the whole network until only one process remains.
The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.
Fig.~\ref{fig:Fusion:Types} shows the type definitions for fusion.
Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
The fused label contains the pair of both original labels, as well as the static part of the $\InputState$ for each input channel.
If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, but rather than knowing its actual value as in the dynamic evaluation, we know its value is stored in the $@chan@$ variable for that channel.

>>>>

% -----------------------------------------------------------------------------
\section{Fusion}
\label{s:Fusion}

The core fusion algorithm works by constructing a static schedule for a single pair of processes. To fuse a whole process group we fuse pairs until only one process remains.

Fig.~\ref{fig:Fusion:Types} shows the type definitions for fusion.
Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
The fused label contains the pair of both original labels, as well as the static part of the $\InputState$ for each input channel.
If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, but rather than knowing its actual value as in the dynamic evaluation, we know its value is stored in the $@chan@$ variable for that channel.

### 2017-02-27 2017mergingmerges/section/XX-Evaluation.tex
Shift discussion of fusion order to Fusion In Practice section, and elaborate

\section{Evaluation}
\label{s:Evaluation}

@@

\subsection{Optimisation}
\label{s:Optimisation}
After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.

>>>>

% -----------------------------------------------------------------------------
\section{Fusion in Practice}
\label{s:Evaluation}

Stream fusion is ultimately performed for practical reasons. We want the fused result program to run faster than unfused source program. Fused result programs like the one in Fig. \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell (maybe using Template Haskell), C or LLVM abstract assembly. The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.

% -----------------------------------------------------------------------------
\subsection{Fusability}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider our @alternatives@ example from \S\ref{s:EvaluationOrder}, here it is again:
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

In our current system, if fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@. As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.

Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstrem @zip@ process then the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which of the instructions from either process can be executed next, and when execution is deadlocked there are none.

On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused. The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels. This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and may other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@. When the dataflow network has a single output stream then we use the method of starting from the process closes to the output stream, walking to towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels where need to be synchronized. In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.

% We discuss other possible approaches in \S\ref{s:Future:FusionOrder}.

% When the two @alt2@ processes are fused we end up with a sequential result process. Trying to then fuse that result into the downstream @zip@ process fails. Alternately, if we were to try to concurrently evaluate the fused result process with the downstream @zip@ process would deadlock because we would reach a state where neither process could advance. To state this again: fusion fails when concurrent evaluation of the two source processes would deadlock.

% The main fusion algorithm here works on pairs of processes.
% When there are more than two processes, there are multiple orders in which the pairs of processes can be fused. The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read. Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process. In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes. The simplest example of this has two input streams, a function applied to both, then zipped together.

% -----------------------------------------------------------------------------
\subsection{Result Size}

@@

% -----------------------------------------------------------------------------
\subsection{Optimisation}
\label{s:Optimisation}
After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.

### 2017-02-27 2017mergingmerges/section/XX-FutureWork.tex
Wibble pagination to see how much space we have left
Shift discussion of fusion order to Fusion In Practice section, and elaborate

%!TEX root = ../Main.tex
\section{Future work}
\label{s:FutureWork}

@@

\subsection{Finite streams}
\label{s:Finite}

@@

\subsection{Non-determinism and fusion order}
\label{s:FusionOrder}

The main fusion algorithm here works on pairs of processes.
When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read.
Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process.
In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes.
The simplest example of this has two input streams, a function applied to both, then zipped together.

\begin{code}
zipMap as bs =
  let as' = map (+1) as
      bs' = map (+1) bs
      abs = zip as' bs'
  in  abs
\end{code}

There are three combinators here, so after converting each combinator to its process there are three orders we can fuse.
The two main options are to fuse the two maps together and then add the zip, or to fuse the zip with one of the maps, then add the other map.
If we start by fusing the zip with one of its maps, the zip ensures that its inputs are produced in lock-step pairs, and then adding the other map will succeed.
However if we try to fuse the two maps together, there are many possible interleavings: the fused program could read all of @as'@ first; it could read all of @bs'@ first; it read the two in lock-step pairs; or any combination of these.
When the zip is added, fusion will fail if the wrong interleaving was chosen.

@@

Our current solution to this is to try all permutations of fusing processes and use the first one that succeeds.
A more principled solution may be to allow non-determinism in a single process by adding a non-deterministic choice instruction.
Then when fusing two processes together, if both processes are pulling from unrelated streams, the result would be a non-deterministic choice between pulling from the first process and executing the first, or pulling from the second process and executing the second.
In this way we could defer committing to a particular evaluation order until the required order is known.
This may produce larger intermediate programs, but the same deterministic program could be extracted at the end.

>>>>

%!TEX root = ../Main.tex

% -----------------------------------------------------------------------------
\section{Future work}
\label{s:FutureWork}

@@

% -----------------------------------------------------------------------------
\subsection{Finite streams}
\label{s:Finite}

@@

% -----------------------------------------------------------------------------
% \subsection{Non-determinism and fusion order}
% \label{s:FusionOrder}
% The main fusion algorithm here works on pairs of processes.
% When there are more than two processes, there are multiple orders in which the pairs of processes can be fused. The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read. Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process. In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes. The simplest example of this has two input streams, a function applied to both, then zipped together. 
% \begin{code}
% zipMap as bs =
%   let as' = map (+1) as
%       bs' = map (+1) bs
%       abs = zip as' bs'
%   in  abs
% \end{code}
% There are three combinators here, so after converting each combinator to its process there are three orders we can fuse. The two main options are to fuse the two maps together and then add the zip, or to fuse the zip with one of the maps, then add the other map. If we start by fusing the zip with one of its maps, the zip ensures that its inputs are produced in lock-step pairs, and then adding the other map will succeed. However if we try to fuse the two maps together, there are many possible interleavings: the fused program could read all of @as'@ first; it could read all of @bs'@ first; it read the two in lock-step pairs; or any combination of these. When the zip is added, fusion will fail if the wrong interleaving was chosen.

@@

% \subsection{Non-determinism and fusion order}
% \label{s:Future:FusionOrder}
% Our current solution to this is to try all permutations of fusing processes and use the first one that succeeds. A more principled solution may be to allow non-determinism in a single process by adding a non-deterministic choice instruction. Then when fusing two processes together, if both processes are pulling from unrelated streams, the result would be a non-deterministic choice between pulling from the first process and executing the first, or pulling from the second process and executing the second. In this way we could defer committing to a particular evaluation order until the required order is known. This may produce larger intermediate programs, but the same deterministic program could be extracted at the end.
